---
title: "GOV-51-Final-Project"
author: "Jenny Jia"
date: "2026-05-11"
output: pdf_document
---

GitHub Repository: https://github.com/jennyjia319/Final-Project
```{r}
#| echo: false # hide unncessary outputs
#| warning: false
#| message: false
```


```{r}
library(knitr)
library(tidyverse)
library(glmnet)
data <- read.csv("data/GOV 51 Master Dataset.csv")
```

```{r}
head(data) # see if csv loaded properly
```

```{r}
tract_data <- data |> # ACS csv orientation is weird, so I have to reorient columns and rows
  rename(variable = 1) |>
  pivot_longer(
    cols = -variable,
    names_to = "tract",
    values_to = "value"
  ) |>
  mutate(
    variable = str_squish(variable),
    value = parse_number(as.character(value))
  ) |>
  pivot_wider(
    names_from = variable,
    values_from = value
  )
```

```{r}
head(tract_data)
```
```{r}
tract_data <- data |> # clean tract names 
  rename(variable = 1) |>
  pivot_longer(
    cols = -variable,
    names_to = "tract",
    values_to = "value"
  ) |>
  mutate(
    variable = str_squish(variable),
    value = parse_number(as.character(value)),
    tract = str_replace_all(tract, "\\.", " "),
    tract = str_remove(tract, " Total Estimate$")
  ) |>
  pivot_wider(
    names_from = variable,
    values_from = value
  ) |>
  select(where(~ !all(is.na(.))))
```
```{r}
head(tract_data) # check to see that it worked 
nrow(tract_data)
colnames(tract_data)
```

# Introduction
Immigration remains one of the most politically contested issues in the United States. Public attitudes toward immigration are often shaped by assumptions about economic contribution, with evidence suggesting that Americans consistently express stronger preferences for highly-skilled immigrants, regardless of their own socioeconomic background (Hainmueller & Hiscox, 2010). What is embedded in these attitudes is the assumption that immigrant presence is closely tied to local economic prosperity.

Existing scholarship has explored the economic consequences of immigration from multiple angles. Peri (2009) finds that immigration can increase productivity and income per worker, while Lin and Weiss (2019) show that these effects are unevenly distributed, with gains concentrated among higher-income groups. Albrizio (2024) further demonstrates that perceptions of immigrant economic contribution often diverge from empirical reality. Yet, despite such substantial literature, most studies focus on estimating immigration’s economic effects rather than asking whether immigrant presence itself can be used as a predictor of neighborhood-level economic conditions. This opens the door to an unexplored puzzle. If immigration shapes local labor markets and economic opportunity, can the share of foreign-born residents help predict neighborhood wealth?

New York City offers a very ideal setting for this question to be explored in because it is one of the most immigrant-dense cities in the United States, with roughly 38% of residents being foreign-born. As well, NYC exhibits substantial variation in both immigrant concentration and median household income across all census tracts. It also contains neighborhoods associated with both high-income, highly educated immigrant communities and lower-income immigrant enclaves, making it a very ideal setting to test predictive relationships.

My project thus asks: Can we predict median household income across census tracts in New York City based on share of foreign-born residents?

This is a predictive observational study, it is not causal. The objective is to assess whether observable demographic characteristics, especially immigrant concentration, can improve tract-level prediction of household income. I find that, while foreign-born share carries predictive information, educational attainment is by far the strongest predictor of neighborhood income.

# Data 
My analysis draws from tract-level data from the 2025 American Community Survey (ACS), combining estimates from the S1903 income dataset and the DP02 selected social characteristics dataset. These sources were merged by census tract identifier to create a unified tract-level dataset for New York City.

The unit of observation is each census tract that is counted within the definition of New York City. After merging and cleaning the ACS data, the dataset contains 2,327 census tracts spanning all five boroughs of NYC: The Bronx, Brooklyn, Manhattan, Queens, and Staten Island.

The outcome variable is median household income. The primary predictor of interest is the percentage of foreign-born residents in each tract. Additional covariates include educational attainment, gender composition, race, age, and total population. Because education is strongly associated with income and may confound the relationship between immigrant concentration and neighborhood wealth, I include a measure of the percentage of residents with at least a bachelor’s degree (so % college-educated includes Bachelor's and Graduate degrees). 

## Data Summary Table
### Table 1: Summary Statistics for NYC Census Tracts

```{r}
data_summary <- tract_data |>
  mutate(
    `College educated (%)` = `Bachelor's degree` + `Graduate or professional degree`
  ) |>
  select(
    `Median income (dollars)`,
    `Foreign-born`,
    `College educated (%)`,
    `Male`,
    `Total population`
  ) |>
  summarise(
    across(
      everything(),
      list(
        Mean = ~mean(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE),
        Min = ~min(.x, na.rm = TRUE),
        Max = ~max(.x, na.rm = TRUE)
      ),
      .names = "{.col}__{.fn}"
    )
  ) |>
  pivot_longer(
    everything(),
    names_to = c("Variable", ".value"),
    names_pattern = "(.+)__(.+)"
  )

data_summary |>
  kable(
    digits = 1,
    caption = "Summary Statistics for NYC Census Tracts"
  )
```
Table 1 shows substantial socioeconomic variation across New York City census tracts, with median household income ranging from roughly 4,930 to 215,581 dollars, highlighting the variation amongst different census tracts in the city. The percentage of foreign-born residents also varies widely, from 0% to 93.3%, while college educational attainment ranges from nearly zero to unanimous attainment, suggesting both may significantly shape local income patterns. By contrast, gender composition is relatively similar across tracts, which means it could serve as a weaker predictor down the line. 

# Analysis 
I begin with descriptive analysis by visualizing the relationship between immigrant concentration and median household income across census tracts. This provides an initial assessment of the direction, strength, and heterogeneity of the possible association.

I then create and use an ordinary least squares (OLS) regression model as an interpretative predictive baseline. Even though my research question is predictive, OLS allows straightforward preliminary interpretation of how tract-level characteristics are associated with income while controlling for potential confounders.

Since predictiveness is my main objective, I also employ three predict tools, those being LASSO, Ridge, and Elastic Net. These approaches introduce  penalization to reduce overfitting and improve out-of-sample prediction, but differ in how they treat predictors. Model performance is evaluated using an 80/20 train-test split and out-of-sample root mean squared error RMSE.

## Descriptive Visualization
### Figure 1: Scatterplot
```{r}
ggplot(
  tract_data |> filter(!is.na(`Foreign-born`), !is.na(`Median income (dollars)`)),
  aes(x = `Foreign-born`, y = `Median income (dollars)`)
) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "lightblue") +
  labs(
    title = "Foreign-Born and Median Household Income Across NYC Census Tracts",
    x = "Percent Foreign-Born",
    y = "Median Household Income ($)"
  )
```
Figure 1 depicts substantial variation in tract-level income outcomes, along with a negative relationship between immigrant concentration and median household income.

## OLS Model
### Table 2: OLS
```{r}
ols <- lm(
  `Median income (dollars)` ~
    `Foreign-born` +
    `Bachelor's degree` +
    `Graduate or professional degree` +
    `Male` +
    `Median age (years)` +
    `Black or African American` + # White and Hispanic omitted avoid multicollinearity among racial composition percentages
    `Asian`,
  data = tract_data
)

summary(ols)
```
In the OLS model depicted in Table 2, the coefficient on foreign-born share is negative but not statistically significant. What this suggests is that the negative relationship between immigrant concentration and income we found earlier is partly explained by other tract-level characteristics, especially educational attainment. On the other hand, education remains a strong predictor, where a one percentage point increase in bachelor’s degree share is associated with approximately 766 dollars higher median household income, while a one percentage point increase in graduate/professional degree share is associated with approximately 1,214 dollar higher median household income, holding other variables constant.

## Penalized Predictive Models 
While OLS provides an easily interpretable benchmark, predictive modeling prioritizes out-of-sample performance that can help reduce overfitting. So, they would therefore better suited for evaluating predictive accuracy. I compare LASSO, Ridge, and Elastic Net to assess whether penalization improves tract-level income prediction relative to the OLS baseline.

### 1. Create Modelling Dataset
```{r}
model_data <- tract_data |>
  mutate(
    college_educated =
      `Bachelor's degree` + `Graduate or professional degree`
  ) |>
  filter(
    !is.na(`Median income (dollars)`),
    !is.na(`Foreign-born`),
    !is.na(college_educated),
    !is.na(Male),
    !is.na(`Median age (years)`),
    !is.na(`Black or African American`),
    !is.na(Asian)
  )
```

### 2. 80/20 Train/test Split 
```{r}
set.seed(51)

train_index <- sample(
  1:nrow(model_data),
  0.8 * nrow(model_data)
)

train <- model_data[train_index, ]
test <- model_data[-train_index, ]
```

### 3. Design
```{r}
X_train <- model.matrix(
  `Median income (dollars)` ~
    `Foreign-born` +
    college_educated +
    Male +
    `Median age (years)` +
    `Black or African American` +
    Asian,
  data = train
)[, -1]

y_train <- train$`Median income (dollars)`

X_test <- model.matrix(
  `Median income (dollars)` ~
    `Foreign-born` +
    college_educated +
    Male +
    `Median age (years)` +
    `Black or African American` +
    Asian,
  data = test
)[, -1]

y_test <- test$`Median income (dollars)`
```

### 4. LASSO
```{r}
lasso <- cv.glmnet(
  X_train,
  y_train,
  alpha = 1
)

lasso_predicted <- predict(
  lasso,
  newx = X_test,
  s = "lambda.min"
)

rmse_lasso <- sqrt(
  mean((lasso_predicted - y_test)^2)
)
```

### 5. Ridge
```{r}
ridge <- cv.glmnet(
  X_train,
  y_train,
  alpha = 0
)

ridge_predicted <- predict(
  ridge,
  newx = X_test,
  s = "lambda.min"
)

rmse_ridge <- sqrt(
  mean((ridge_predicted - y_test)^2)
)
```

### 6. Elastic Net
```{r}
elastic <- cv.glmnet(
  X_train,
  y_train,
  alpha = 0.5
)

elastic_predicted <- predict(
  elastic,
  newx = X_test,
  s = "lambda.min"
)

rmse_elastic <- sqrt(
  mean((elastic_predicted - y_test)^2)
)
```

### 7. OLS RMSE
```{r}
ols_predicted <- predict(
  ols,
  newdata = test
)

rmse_ols <- sqrt(
  mean((ols_predicted - y_test)^2)
)
```

### 8. Table 3: Comparison Table
```{r}
comparison_table <- tibble(
  Model = c(
    "OLS (no regularization)",
    "LASSO (alpha = 1)",
    "Ridge (alpha = 0)",
    "Elastic Net (alpha = 0.5)"
  ),
  `Test RMSE` = c(
    rmse_ols,
    rmse_lasso,
    rmse_ridge,
    rmse_elastic
  )
)

comparison_table |>
  kable(
    digits = 1,
    caption = "Out-of-sample Prediction Performance: OLS vs Penalized Approaches"
  )
```
Table 3 compares out-of-sample predictive performance in both OLS and penalized predictive approaches. Surprisingly, what we find is that the OLS model produces the lowest test RMSE (14,591.5), outperforming LASSO (14,798.3), Elastic Net (14,797.4), and Ridge (14,907.6). This suggests that regularization does not improve predictive accuracy in this case, mostly likely because the model includes a relatively small number of relevant predictors and does not appear to suffer from severe overfitting or noise. 

### 9. Figure 2: Comparing Penalized Approaches
```{r}
plot(log(lasso$lambda), lasso$cvm, type = "l",
     xlab = "log(Lambda)",
     ylab = "Cross-Validated MSE",
     col = "lightblue")

lines(log(ridge$lambda), ridge$cvm, col = "darkblue")
lines(log(elastic$lambda), elastic$cvm, col = "lightpink")

legend("topright",
       legend = c("LASSO", "Ridge", "Elastic Net"),
       col = c("lightblue", "darkblue", "lightpink"),
       lty = 1)
```
Figure 2 compares MSE across LASSO, Ridge, and Elastic net models as the  penalty increases. At lower penalty values, all three models exhibit very similar predictive performance, suggesting that penalization has limited effect when coefficient shrinkage is modest. As the penalty increases, however, prediction error rises sharply, especially for LASSO and elastic net. This means  that excessive regularization removes too much predictive power and worsens model performance. Ridge remains relatively more stable under stronger penalization, though Table 3 from above shows that none of the penalized approaches actually outperform the simpler OLS benchmark in out-of-sample prediction.

# Conclusionn & Limitations
## Conclusion
My study investigated whether immigrant concentration serves as a meaningful predictor of median household income across New York City census tracts. Descriptive analysis initially pointed to a negative relationship between foreign-born share and neighborhood income, however, this association dimmed considerably once tract-level socioeconomic and demographic controls were introduced. Within the expanded OLS model, immigrant concentration lost its statistical significance as an independent predictor, while educational attainment emerged as the dominant driver of income variation at the tract level.

Thinking about this predictively, my results point to the fact that, rather than immigrant concentration alone, is what drives model performance. Foreign-born share carries some descriptive signal, but most of the predictive power stems from educational attainment and the demographic makeup of a Census tract. Also, very interestingly and surprisingly, the more complex penalized  approaches failed to surpass the simpler OLS benchmark, suggesting the model's predictive signal was already stable enough that coefficient shrinkage offered minimal additional value.

Overall, these findings are largely surprising, especially given the often-concieved notion that urban immigrants tend to take higher-income jobs. If anything, my results suggest that this narrative may be more of a myth than reality, at least at the neighborhood level in NYC. It is possible that, while some immigrant groups do enter high-wage sectors, the tract-level indication of that fact is diluted by the diversity within the foreign-born population itself, which might mask any high income that might exist at the individual level.

## Limitations 
In addressing limitations, my analysis sought to be observational in nature, so no causal interpretation of immigration's role in shaping neighborhood income can be concluded from this study. My model also omits potentially consequential predictors, for example, occupation and different industries. That being said, such specific data is unavailable and is not represented in ACS datasets. Finally, my focus on New York City limits the generalizability of these findings, as other metropolitan areas with distinct immigration and income profiles may yield different results.

# Works Cited
Albrizio, S. (2024). Shifting perceptions: Unpacking public support for immigrant workers integration in the labor market. IMF Working Paper, 2024(217), 1. https://doi.org/10.5089/9798400290374.001

Center for Migration Studies. (2025). Data Briefing: A Portrait of Immigrant New Yorkers. https://cmsny.org/publications/data-briefing-on-new-york-city-immigrants/

Hainmueller, J., & Hiscox, M. J. (2010). Attitudes toward Highly Skilled and Low-skilled Immigration: Evidence from a Survey Experiment. American Political Science Review, 104(1), 61–84. https://doi.org/10.1017/s0003055409990372

Lin, K., & Weiss, I. (2019). Immigration and the wage distribution in the United States. Demography, 56(6), 2229–2252. https://doi.org/10.1007/s13524-019-00828-9

Peri, G. (2009). The Effect of Immigration on Productivity: Evidence from US States. In National Bureau of Economic Research. https://doi.org/10.3386/w15507